Machine Learning Tools for Open Cluster Characterization with Gaia DR2 Data

### 2021-02-07 paper/paper.tex
paper: Update Related works
Update Related works section with @cguz suggestions.

Co-Authored-By: Cesar Guzmann <15159632+cguz@users.noreply.github.com>

The initial approach to find open clusters is by searching for overdensities
in the astrometric space of the galactic disk and a later identification of
possible OCs using photometric information \cite{castro2020hunting}.
However, although it looks simple, it is not. Here we face the problem that the
near field around the OC is composed of two kind of populations.
The first kind belongs to the membership stars of the OC (from tens or hundreds
to a few thousand) while the second one is composed by a background of stars
that do not belong to the OC (from tens to hundreds of thousands).
To find out which stars do belong to the OC is the problem we try to solve in
this work as this selection is critical to properly characterize the
fundamental properties of the cluster (total mass, age, chemical composition, etc.).

Since the components of the open cluster are born from the same gas cloud and they
share both dynamic properties and relative distance, as well as age and chemical
composition, the first step is to analyze the proper motion configuration space of
the region of interest. This approach is the one used by many tools from the VO
such as TOPCAT \cite{taylor2005topcat}, Clusterix 2.0 \cite{balaguer2020clusterix},
Aladin \cite{bonnarel2000aladin} or VOSA \cite{bayo2008vosa}.

TOPCAT by itself cannot find members of the open cluster. It requires to parameterize
some of the properties of the cluster, what requires a previous knowledge of the cluster.
Or to make a supervised and manual selection of groups based on overdensities
in the proper motion configuration space. Clusterix 2.0, an interactive web-based tool,
can help to this process. It takes the proper motion diagram without making any prior
assumption about the membership of the candidate star and determines empirically
the frequency functions. Clusterix uses normal Gaussian kernel functions, defined as:

\begin{equation*}
  K(a, b) = \frac{1}{2 \pi h^{2}} \exp{ \left[ - \frac{1}{2}\frac{\left( a - a_{i} \right)^{2} + \left( b - b_{i} \right)^{2}}{ h^{2}} \right]}
\end{equation*}

$\left( a, b \right)$ are referred to the proper motion configuration space,
while a point located in the center of the cell $\left( i, j \right)$
provides the maximum contribution for calculating the local density.
Number $h$ is called the \emph{smoothing} parameter and it is measured
in the same units as proper motion.

Clusterix is defined as an unsupervised and non-parameterized tool, but it
critically depends on the selection of three radii in the studied region.
The inner radius should contain stars belonging or not to the cluster.
While the outermost radii define a ring which only contains stars that do
not belong to the cluster. In addition, it is also sensitive to other
parameters, such as the soften parameter \emph{h}, or other restrictions
related with the searched proper motions.

If the analysis performed by Clusterix is successful, it returns the probability
associated to each star in the dataset to belong to the open cluster.
These results can be imported then into TOPCAT to continue the refining process to
get a valid characterization.

Another recent approach to solve this problem based on machine learning techniques
makes a systematic search for overdensities in the astrometric space of the galactic
disk and a subsequent identification of open clusters using photometric information
\cite{castro2020hunting}.

This last method includes two phases: the first one uses an unsupervised clustering
algorithm, DBSCAN, to search for overdensities $(l, b \pi, \mu_{\alpha} *, \mu_{\delta})$,
and then applies a deep learning ANN, previously trained with magnitude diagrams,
to identify isochrone patterns within the detected overdensities and thus proceed
to confirm them as OC.

It should be noted that for the execution of this method, MareNostrum 42
(Barcelona Supercomputing Center) was used.
So the neural network could handle the image recognition process with
isochrone patterns and not applying theoretical models derived from
values such as metallicity or masses, among other.

This work concludes with the identification of 582 new open clusters
distributed along the galactic disk for a galactic declination bellow 20 degrees.
This result increases the number of known open clusters by 45\%.

As told before, our aim in this work, linked to computational limitations compared
to MareNostrum, is not to do a blind search for clusters, but to obtain a method,
in the machine learning environment, that allows the characterization of open clusters.
We require our method to be unsupervised and non-parameterized making it suitable for
automated processes.

We use the Clusterix+TOPCAT method applied to the same clusters that we have analyzed
with our method to validate our results.

In this work we make use of Gaia DR2 since DR3 has not been released in time for us to include it.
Gaia DR2 is a multidimensional dataset obtained by ESA's Gaia mission
(located at L2, 1.5 million kilometers from Earth) and operational since 2014.
The catalogue has high precision and accuracy astrometric data for more than 1.7 billion stellar sources,
and magnitudes in three photometric filters (G, BP and RP) for more than 1,300 million sources.

As we will explain later, what we need to accomplish our aim is an algorithm
that manages to make data groups based on the dynamic properties of the stars.

There are several clustering algorithms: K-Means, Mean-Shift Clustering, DBSCAN \cite{ester1996density}
among other. While each one behaves better according to the distribution of the objects
to clusterize, we have chosen K-Means by its simplicity and good results.

This algorithm works well and gives good results at first approach.
However, we need to set a large number of clusters to find the open cluster we are looking for.
This fact complicates the identification of the OC and many times the found cluster contains to many outliers.
For that reason, we have search for a K-Means refinement based on an artificial neural network.

The \emph{Unsupervised Deep Embedding for Clustering Analysis} model
or \emph{DEC} \cite{xie2016unsupervised} takes K-Means as its starting point,
but then, it trains an autoencoder to reduce the feature space
and pass this transformed data through a Clustering Layer which refines the previous selection.

>>>>

In~\cite{castro2020hunting}, they present one of the initial approach to detect OCs.
They search for overdensities in the astrometric space of the galactic disk, and then,
they identify with photometric information possible OCs. Although it looks simple,
it is quite difficult to face the problem that the near field around the OC has two
kind of populations. The first population belongs to the membership stars of the OC
(from tens or hundreds to a few thousands), while the second population has a background
of stars that do not belong to the OC (from tens to hundreds of thousands). Thus,
we summarize our problem to find out which stars belong to the OC. This is a handicap
work because the selection is critical to properly characterize the fundamental
properties of the cluster (total mass, age, chemical composition, between others).

Other works, such as TOPCAT~\cite{taylor2005topcat}, Clusterix 2.0~\cite{balaguer2020clusterix},
Aladin~\cite{bonnarel2000aladin} or VOSA~\cite{bayo2008vosa}, perform as a first step
to analyze the proper motion configuration space of the region of interest.

TOPCAT cannot find open cluster, instead it requires to parameterize some of the
properties of the cluster. This last process requires a previous knowledge of the cluster.
In other words, they perform a supervised and manual selection of groups based on
overdensities in the proper motion configuration space. Clusterix 2.0 is an interactive
web-based tool that can help to the selection of groups. It takes the proper motion
diagram without making any prior assumption about the membership of the candidate star
and determines empirically the frequency functions. It employs normal Gaussian kernel
functions, defined as:

\begin{equation*}
  K(a, b) = \frac{1}{2 \pi h^{2}} \exp{ \left[ - \frac{1}{2}\frac{\left( a - a_{i} \right)^{2} + \left( b - b_{j} \right)^{2}}{ h^{2}} \right]}
\end{equation*}

where \(( a, b )\) are the proper motion configuration space, \(( i, j )\) is a point
located in the center of that provides the maximum contribution for calculating the
local density, and \(h\) is the \emph{smoothing} parameter, which it is measured in the
same units as the proper motion.

Clusterix critically depends on the selection of three radii in the studied region.
The inner radius contains stars belonging or not to the cluster. While the outermost
radii defines a ring which only contains stars that do not belong to the cluster.
In addition, it is also sensitive to other parameters, such as the soften parameter
\emph{h}, or other restrictions related with the searched proper motions.

If the analysis performed by Clusterix is successful, it returns the probability of
each star to belong to the open cluster. The results can be imported then into TOPCAT
to continue the refining process to get a valid characterization.

Another recent approach~\cite{castro2020hunting} solves the problem using machine
learning techniques. It searches systematically for overdensities in the astrometric
space of the galactic disk and a subsequent identification of open clusters using
photometric information. It includes two phases:

\begin{enumerate}
  \item it employs DBSCAN, an unsupervised clustering algorithm, to search for
        overdensities,
  \item and it applies a deep learning ANN to identify isochrone patterns within
        the detected overdensities and thus proceed to confirm them as OCs. They
        trained previously the ANN with magnitude diagrams.
\end{enumerate}

They execute their experiments in the Barcelona Supercomputing Center, MareNostrum 42.
Thereby, the neural network handles the image recognition process with isochrone
patterns without applying theoretical models derived from values such as metallicity
or masses, among other. Their work concludes with the identification of 582 new open
clusters distributed along the galactic disk for a galactic declination bellow 20
degrees, increasing the number of known open clusters by 45\%.

In contrast to~\cite{castro2020hunting}, we lack of a super computer like MareNostrum
for our experiments. For this reason, our aim is to obtain a new novel method that
allows the characterization of open clusters without doing a blind search for clusters
characterization. We advocate of the idea of building an unsupervised and
non-parameterized method. So, it an be suitable for automated processes.
